### 2024-03-06 chapter_1/src/manuscript.qmd

In this chapter I will present a theory of how political visual objects are used by the public to make decisions. As I will argue throughout the chapter, existing models of political information processing set the foundation for the theory I present here. However, there requires adjustment to the underlying cognitive mechanisms to account for differences in how the brain processes visual information relative to things like a stated policy position of a political candidate that you read in the news paper or hear while watching a video.

@@

Why is it necessary for me to start with simple cases? Visual information, at large, is comparatively understudied in the field. The field largely considers information that are quite common and explicitly political such as stated policy preferences, conversations with our friends about politics, and text with political information from the news or campaigns. This is not an indictment on the field. In fact, political psychologists have made significant strides to theorize about political information processing. The dominant theories coming from these interdisciplinary efforts are at the core of the theory that I present here. The reason there is relatively little theorizing about visual information and not more explicitly political information is likely due to the methodological challenges that come with theorizing about visual experiences.

@@

Before presenting a theory of visual information processing in politics, I will first discuss the dominant models of political information processing. As I have mentioned, the mechanisms outlined in these theories are at the core of the theory that I present. Both practically and conceptually, my model of visual information processing amends these existing models so that they may better generate predictions about how visual information influences political cognition. After introducing some of the key mechanisms outlined in those theories, I will present the theory that is the focus of the book. I will refer to this theory as the snap-judgment model of political information processing. Once I explain this theory, I will close this chapter with a discussion about how this contrasts with existing theories of political information processing, but I will more importantly discuss how this theory contributes to these existing models.

@@

There are two dominant models of how the public in the United States process political information and express attitudes in response to such information. There are, of course, many more than these two; particularly in the fields of survey and public opinion research [see @tourangeau_psychology_2000]. I will focus on the two most commonly used in political science, however, as they synthesize many of these other alternative models and are quite commonly used by political psychologists.

@@

In the "memory" model, @zaller_nature_1992 characterizes the public as Bayesian updaters who, when gathering information about politics, will develop a cognitive schema about how that information relates to existing information. That is, citizens are not passively consuming political information from political elites -- which is much of the focus in @zaller_nature_1992. Rather, the public actively integrate this information. Nevertheless, they do not weigh all new information the same. The public have a number of "predispositions" that have influence on the degree to which and *how* we organize this new information. As a result of these features, not all new political information will prove influential in our expressed political attitudes all of the time.

@@

The next canonical model of political information processing is Lodge and Taber's [-@lodge_rationalizing_2013] "online" model. This model also views the public as selectively incorporating Bayesian updaters. Like Zaller's [-@zaller_nature_1992] "memory" model, the Lodge and Taber's [-@lodge_rationalizing_2013] "online" model is heavily interdisciplinary. However, they characterize this "Receive" and "Accept" processes of political information as slightly different with significant implications.

One of the psychological theories that Lodge and Taber incorporate into their model of psychological processing is referred to as "motivated reasoning" [see @taber_motivated_2006]. The psychological theory of motivated reasoning suggests that the degree to which we gather new information that goes against our pre-existing viewpoints depends on which motivation we presently have. In summarizing this theory, @kunda_case_1990 argues that in contexts where we are motivated to gather information so that we are accurate we are more likely to incorporate all new information; even new information that challenges our pre-existing views. However, according to @kunda_case_1990, we often default to the motivation of selectively gathering information that fits with our pre-existing views in an attempt to have more evidence that we can cite to defend our position to those who may wish to challenge them. This theory of motivated reasoning has been and is still quite popular among social psychologists who examine how members of social groups might justify their negative views of those in another group while maintaining positive views of those in their own group, despite the presence any contradictory evidence [see @redlawsk_affective_2010].

@@

Let me first start off by departing from talking about politics for a minute and consider a related topic: marketing and the role of color. Visual branding is a significant factor in predicting consumer behavior. Marketing is not simply producing catchy slogans but is also about how to visually present a brand. There is ample evidence suggesting that choices in visual presentation has significant influence on whether or not a consumer will purchase one of your products. For example, evidence suggests that hotels that have colors like red are much less likely to be booked relative to hotel rooms that feature softer colors like blues and beiges [@kuo_impact_2023].

As this example demonstrates, color is a significant form of visual information that can influence how we behave. But, why? There is a rich literature in marketing referred to as color theory which suggests that colors evokes particular affective and emotive reactions. For example, @cimbalo_emotionally_1978 demonstrates that colors like red and blue contain different emotive associations. Specifically that red can evoke negative emotions and blue can contain positive and more soothing emotions.

Connecting this idea to that of the work that conceptualizes memory as an associative network, colors are not simply just something that exists but it is a node that contains its own associations. Like any node or memory, color does not have to hold one connection nor do all connections have equal strength under every circumstance. Let me illustrate this with a simple example.

@@

@fig-color-associations presents all the different things that we may associate the color yellow with. We might associate it with our favorite sweatshirt; with a food item, a banana; the sunrise; and the Los Angeles Rams. You may think of different things when you think of the color yellow. This is because you may have different connections to the "yellow node" in your head than I do. This arises with different experiences and differences in the salience of those associations because of your cognitive schema. However, 3 months from now you may think of different things when asked what you associate with the color yellow. This is because your schema is reorganized or different aspects of your schema are accessed because of the context you are in. Perhaps, today you think of the Los Angeles Rams, but say 3 months from now it is April and you are thinking about the Milwaukee Brewers because it is now baseball season and the Super Bowl happened a couple months ago.

@@

Colors can contain affective associations through connections to the affective associations we have with related objects. Affect is a strong and central part to human memory [see @fazio_attitudes_2007;@kensinger_affective_2022]. When we think back to some event in our life, rarely do we only recall what happened but we often recall how it made us feel -- and sometimes we may start to feel happy, angry, or fearful like we did in that recalled event. This is because memories are simply refired electrical signals between a particular network of neurons that were originally active during experiencing (encoding) the event [@fazio_attitudes_2007]. Going back to the idea of affect transfer, as discussed by Lodge and Taber [-@lodge_rationalizing_2013] who applied in in the context of politics, colors contain affective information not necessarily as a result of any inherent property about the color but because of the affect that comes with the objects that we commonly associate with that color [@cimbalo_emotionally_1978].

@@

In the context of the United States, the colors red and blue are ubiquitous in politics. Since the 2000 Presidential Election, news organizations have consistently converged on the use of the colors red and blue to convey to the public whether the Republican or the Democrat, respectively, won a particular electoral district [@elving_color_2014]. Extending these ideas about how associations between color and some object can come to form in the brain, we may expect that the associations of the colors red and blue with Republicans and Democrats in the United States would occur.

@@

First, we receive visual information as wavelengths that are passed to our retina. Our retina passes this information to our optic nerve as an electrical signal to our visual cortex. Once in our visual cortex, the process of connecting those nodes of visual information in the cortex activates the paths to related nodes across our brain [@goldstein_sensation_2017]. While this glosses over important physiological components such as the role of ganglion cells and the differences between the V1 and V4 sections of the visual cortex, this still illustrates how the information is received and starts those now familiar cognitive processes.

Once the visual information has been decoded, our brain will want to find information, or meaning, from those electrical signals. It therefore starts looking down all of these forking paths that connect nodes together. In other words, the associative network has been activated. Once activated, we will continue to spend effort on paths that continually provides more relevant information and will abandon paths that are less relevant and informative.

@@

So, how does this work with political visual information? @fig-theory presents what I call the snap-judgment model of politically-relevant visual information processing.

@@

Once one consumes visual information about politics, this cognitive process that I have belabored is activated. One will seek to find political meaning from that visual information. This pre-conscious process that the brain starts will go down these finite forking paths to find which associated node provides information about the current stimulus.

>>>>

In this chapter I will present a theory of how political visual objects are used by the public to make decisions. *That is, politically-relevant visual objects are information for the public.* I consider existing models of political information processing as the foundation for the theory I present here. Nevertheless they require adjustment to account for the underlying cognitive mechanisms to account for differences in how the brain processes visual information relative to things like a stated policy position of a political candidate that you read in the news paper or hear while watching a video.

@@

Why is it necessary for me to start with simple cases? Visual information, at large, is comparatively understudied in political science. The field largely considers information that are quite common and explicitly political such as stated policy preferences, conversations with our friends about politics, and text with political information from the news or campaigns. This is not an indictment on the field. In fact, political psychologists have made significant strides to theorize about political information processing. The dominant theories coming from these interdisciplinary efforts are at the core of the theory that I present here. The reason there is relatively little theorizing about visual information and not more explicitly political information is likely due to the methodological challenges that come with theorizing about visual experiences.

@@

Before presenting a theory of visual information processing in politics, I will first discuss the dominant models of political information processing. As I have mentioned, the mechanisms outlined in these theories are at the core of the theory that I present. Both practically and conceptually, my model of visual information processing amends these existing models so that they may better generate predictions about how visual information influences political cognition. After introducing some of the key mechanisms outlined in those theories, I will present the theory that is the focus of the book. I will refer to this theory as *the snap-judgment model of political information processing*. Once I explain this theory, I will close this chapter with a discussion about how this contrasts with existing theories of political information processing, but I will more importantly discuss how this theory contributes to these existing models.

@@

There are two dominant models of how the public in the United States process political information and express attitudes in response to such information. There are, of course, many more than these two; particularly in the fields of survey and public opinion research [see @tourangeau_psychology_2000]. I will focus on the two most commonly used in political science, however, as they synthesize many of these other alternative models and are quite commonly used by political psychologists due to their focus on mechanisms about how information is encoded and converted into an attitude.

@@

In the "memory" model, @zaller_nature_1992 characterizes the public as Bayesian updaters who, when gathering information about politics, will develop a cognitive schema about how that information relates to existing information. That is, citizens are not passively consuming political information from political elites. Rather, the public actively integrate this information. Nevertheless, they do not weigh all new information the same. The public have a number of "predispositions" that have influence on the degree to which and *how* we organize this new information. As a result of these features, *not all new political information will prove influential in our expressed political attitudes all of the time*.

@@

The next canonical model of political information processing is Lodge and Taber's [-@lodge_rationalizing_2013] "online" model. This model also views the public as selectively incorporating Bayesian updaters. Like Zaller's [-@zaller_nature_1992] "memory" model, the Lodge and Taber's [-@lodge_rationalizing_2013] "online" model is heavily interdisciplinary. However, they characterize this "Receive" and "Accept" processes of political information as slightly different mechanistically with significant implications.

One of the psychological theories that Lodge and Taber incorporate into their model of psychological processing is referred to as "motivated reasoning" [see @taber_motivated_2006]. The psychological theory of motivated reasoning suggests that the degree to which we gather new information that goes against our pre-existing viewpoints depends on which motivation we presently have. In summarizing this theory, @kunda_case_1990 argues that in contexts where we are motivated to gather information so that we are accurate we are more likely to incorporate all new information; even new information that challenges our pre-existing views. However, according to @kunda_case_1990, we often default to the motivation of selectively gathering information that fits with our pre-existing views in an attempt to have more evidence that we can cite to defend our position to those who may wish to challenge them. This theory of motivated reasoning has been and is still quite popular among social psychologists who examine how members of social groups might justify their negative views of those in another group while maintaining positive views of those in their own group, despite the presence of contradictory evidence [see @redlawsk_affective_2010].

@@

Let me first start off by departing from talking about politics for a minute and consider a related topic: marketing and the role of color. Visual branding is a significant factor in predicting consumer behavior. Marketing is not simply producing catchy slogans but is also about how to visually present a brand. There is ample evidence suggesting that choices in visual presentation has significant influence on whether or not a consumer will purchase a product. For example, evidence suggests that hotels that have colors like red are much less likely to be booked relative to hotel rooms that feature softer colors like blues and beiges [@kuo_impact_2023].

As this example demonstrates, color is a significant form of visual information that can influence how we behave. But, why? There is a rich literature in marketing referred to as color theory which suggests that colors evokes particular affective and emotive reactions. For example, @cimbalo_emotionally_1978 demonstrates that colors like red and blue contain different emotive associations. Specifically, @cimbalo_emotionally_1978 finds evidence suggesting that red can evoke negative emotions and blue can contain positive and more soothing emotions.

Connecting this idea to that of the work that conceptualizes memory as an associative network, colors are not simply just something that exists but it is a node that contains its own associations. Like any node or memory, color does not have to hold one connection nor do all connections have equal strength under every circumstance. That is, while red has a tendency to evoke negative emotions in marketing settings, it does not necessarily mean it is strictly emotively negative under every circumstance due to its potential to be connected to nodes with positive associations. Let me illustrate this with a simple example.

@@

@fig-color-associations presents all the different things that we may associate the color yellow with. We might associate it with our favorite clothing item, a sweatshirt; with a food item, a banana; a time of day, the sunrise; and an American football team, the Los Angeles Rams. You may have come up with a different list. This is because you may have different connections to the "yellow node" in your head than I do. This arises with different experiences and differences in the salience of those associations because of your cognitive schema. However, 3 months from now you may think of different things when asked what you associate with the color yellow. This is because your schema is reorganized or different aspects of your schema are accessed because of the context you are in. Perhaps, today you think of the Los Angeles Rams, but say 3 months from now it is April and you are thinking about the Milwaukee Brewers because it is now baseball season and the Super Bowl happened a couple months ago.

@@

Colors can contain affective associations through connections to the affective associations we have with related objects. That is, affect can transfer due to intermediary associations. Affect is a strong and central part to human memory [see @fazio_attitudes_2007;@kensinger_affective_2022]. When we think back to some event in our life, rarely do we only recall what happened but we often recall how it made us feel -- and sometimes we may start to feel happy, angry, or fearful like we did in that recalled event. This is because these particular connected nodes to constitute a memory are simply refired electrical signals between a particular network of neurons that were originally active during experiencing (encoding) the event [@fazio_attitudes_2007]. Going back to the idea of affect transfer, as discussed by Lodge and Taber [-@lodge_rationalizing_2013] who applied in in the context of politics, colors contain affective information not necessarily as a result of any inherent property about the color but because of the affect that comes with the objects that we commonly associate with that color [@cimbalo_emotionally_1978].

@@

In the context of politics in the United States, the colors red and blue are ubiquitous. Since the 2000 Presidential Election, news organizations have consistently converged on the use of the colors red and blue to convey to the public whether the Republican or the Democrat, respectively, won a particular electoral district [@elving_color_2014]. Extending these ideas about how associations between color and some object can come to form in the brain, we may expect that the associations of the colors red and blue with Republicans and Democrats in the United States would occur.

@@

First, we receive visual information as wavelengths that are passed to our retina [@goldstein_sensation_2017]. Our retina passes this information to our optic nerve as an electrical signal to our visual cortex [@goldstein_sensation_2017]. Once in our visual cortex, the process of connecting those nodes of visual information in the cortex activates the paths to related nodes across our brain [@goldstein_sensation_2017]. While this glosses over important physiological components such as the role of ganglion cells and the differences between the V1 and V4 sections of the visual cortex [see @goldstein_sensation_2017], this still illustrates how the information is received and starts those now familiar cognitive processes.

Once the visual information has been decoded, our brain will want to find information, or meaning, from those electrical signals. It therefore starts looking down all of these forking paths that connect nodes together. In other words, the associative network has been activated. Once activated, we will continue to exert effort on paths that continually provides more relevant information and will abandon paths that are less relevant and informative.

@@

So, how does this work with political visual information? @fig-theory presents what I call the snap-judgment model of politically-relevant visual information processing. Note, in the following chapters I will go over this process again specific to the political context of topic to demonstrate the contextual flexibility of this model. Here, I will provide a very general overview of the model.

@@

Once one consumes visual information about politics, this cognitive process of associative networks that I have belabored is activated. One will seek to find political meaning from that visual information. We do not have to tell our brain to find meaning from this information, it just does it automatically as part of its pre-conscious processing. This pre-conscious process that the brain starts will go down these finite forking paths to find which associated node provides information about the current stimulus.
